\propertitle

### 2019-06-18 maxNt_v5.tex
update

\pdfoutput=1
%% Author: PGL  Porta Mana
%% Created: 2015-05-01T20:53:34+0200
%% Last-Updated: 2019-06-17T17:13:33+0200
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newif\ifarxiv
\arxivfalse
\ifarxiv\pdfmapfile{+classico.map}\fi
\newif\ifafour
\afourfalse% true = A4, false = A5
\newif\iftypodisclaim % typographical disclaim on the side
\typodisclaimtrue
\newcommand*{\memfontfamily}{zplx}
\newcommand*{\memfontpack}{newpxtext}
\documentclass[\ifafour a4paper,12pt,\else a5paper,10pt,\fi%extrafontsizes,%
onecolumn,oneside,article,%french,italian,german,swedish,latin,
british%
]{memoir}
\newcommand*{\updated}{\today}
\newcommand*{\firstdraft}{4 November 2015}
\newcommand*{\firstpublished}{***}
\newcommand*{\propertitle}{Maximum-entropy distributions for a 
  neuronal network\\ from subnetwork data%\\{\large ***}%
}
\newcommand*{\pdftitle}{\propertitle}
\newcommand*{\headtitle}{Maximum-entropy from sample data}
\newcommand*{\pdfauthor}{P.G.L.  Porta Mana, Y. Roudi, V. Rostami, E. Torre}
\newcommand*{\headauthor}{Porta Mana \etal}
\newcommand*{\reporthead}{}% Report number

@@

\newcommand{\appropto}{\mathrel{\vcenter{
  \offinterlineskip\halign{\hfil$##$\cr
    \propto\cr\noalign{\kern2pt}\sim\cr\noalign{\kern-2pt}}}}}
% https://tex.stackexchange.com/a/33547/97039
\newcommand*{\fm}{f}
\newcommand*{\Fm}{F}

@@

We apply our approach to two concrete data sets: the activity of 65 neurons
recorded from a rat's Medial Entorhinal Cortex \mynote{ref}, and the
activity of 159 neurons recorded from a macaque's Motor Cortex
\mynote{ref}\mynote{add recording length}. In the first data set, the most
interesting inference is that the most probable time-frequency distribution
for the total activity of the full network, if the latter is assumed to
be larger than 1\,000 neurons, has two very distinct modes, both at low
activities ($0.7\%$ and $2.6\%$ of simultaneously active neurons within
$3\;\mathrm{ms}$, see \fig***). The analogous distribution for second data
set doesn't have two modes but still presents one prominent shoulder in its
one low-activity mode. Note that these inferred features for the full
network aren't observed in the sample, nor are they inferred by the
application of maximum-entropy to the sample alone. \mynote{if we manage
  the Monte Carlo sampling or other approx: say something about the
  deviation from the most probable frequency distribution}

@@

Our questions concern general features of the total activity $\yA$ of the
full network during and after the recording. For example: what was its
frequency distribution in time? How much total activity should we expect at
a time bin during the recording, or at a new time bin? We cannot answer
these questions with certainty; we can only give a distribution of
probability over the possible answers. Our approach gives such a
probability distribution.

@@

The idea behind the approach is easily summarized:
\begin{enumerate}[label=(\alph*)]
  \item\label{item:maxent_step} we build a distribution for the total
  activity of the full network using the maximum-entropy method;
  \item\label{item:sample_step} the expectation constraints used in the
  maximum-entropy method for the full network are, in turn, determined
  via sampling theory from the expectation constraints for the sample.
\end{enumerate}
Let's discuss these points in detail.

Regarding~\ref{item:maxent_step}, we assume that you're familiar with the
maximum-entropy method. In short, it amounts to a pair of prescriptions:
choose the distribution having maximum Shannon entropy among those
satisfying specific convex constraints, such as fixed expectation values;
and judge these expectation values to be equal to measured averages. We add
two remarks about this method that are seldom made in the literature.
First, the \enquote{distribution} given by this method is the zeroth-order
approximation of four different distributions:
\begin{itemize}
\item the most probable \emph{time-frequency} distribution of the total
  activity across the \emph{recorded} bins,
\item the most probable \emph{time-frequency} distribution of the total
  activity in a very long run of \emph{new} time bins,
\item the \emph{probability} distribution for the value of the total
  activity at any time bin among those \emph{recorded},
\item the \emph{probability} distribution for the value of the total
  activity at a \emph{new} time bin.
\end{itemize}
The maximum-entropy distribution is thus an approximate answer to four
conceptually different inferences. Note that although the four
distributions above numerically coincide in this approximation, they differ
in higher-order approximations. We discuss their differences in \sect***.
The maximum-entropy distribution approximates best the first of the four
distributions, so that's the one we'll mainly have in mind, and we'll
denote it by $\Fm(\yA)$.***
Second, the maximum-entropy method based on the Shannon entropy implicitly
makes specific assumptions about the probability of the long-run frequency
distributions \citep{jaynes1986d_r1996,portamana2009,portamana2017}. We
discuss these assumptions in \sect***.

In our case, to apply the method for the total activity of the full network
we need to fix some of the latter's expectation values, for example its
moments. But we don't have any measured averages for the full network to
equate our expectations to. Here enters point~\ref{item:sample_step}: the
probability calculus gives an exact, linear relation between the first $m$
moments for the full network and the first $m$ for the sample
\citep[\eqns~(16)]{portamanaetal2015}; the ones determine the others and
vice versa at every time bin. This relation is a classical result of
sampling theory
\citep[\chaps~I--IV]{whitworth1867_r1965}[\chap~II]{feller1950_r1968}[\chap~3]{jaynes1994_r2003}[see
also][]{whitworth1897}.

Combining this result with the maximum-entropy prescription
\enquote{expectations${}={}$measured averages} for the sample, we have that
the measured averages for the sample determine the moments for the full
network:
\begin{equation*}
  \label{eq:chain_prescription_moments}
  \mathrlap{\overbracket{\phantom{\text{measured averages} \rightarrow
        \text{sample moments}}}^{\textit{maximum-entropy prescription}}}
  \text{measured averages} \rightarrow
        \underbracket{\text{sample moments}
  \rightarrow
    \text{full-network moments}}_{\textit{sampling theory}}
\end{equation*}

These two steps are more straightforward if instead of power moments we use
\emph{normalized factorial moments}. The $m$th normalized factorial moment
of the probability distribution for the activity of the recorded neurons is
defined as the expectation
\begin{equation}
  \E\condt[\Bigg]{
    \binom{n \ya}{m}\,\binom{n}{m}^{-1}}{\yH}
  \equiv
  \sum_{\ya} %m!\,
  \binom{n \ya}{m}\,  \binom{n}{m}^{-1}\, \pf(\ya \| \yH),
  \qquad 1\le m \le n
\end{equation}
where $\pf(\ya \| \yH)$ is the probability for activity $\ya$ given
assumptions $\yH$; \enquote{$\ya$} may refer to any recorded bin or to a
new bin. This moment can be interpreted as the expectation of the number of
distinct $m$-tuples of simultaneously spiking neurons (within a bin's time
width), normalized by the number of distinct $m$-tuples. For example, with
$m=2$, if $na=4$ neurons spike in a network of $n=5$, we have
$\tbinom{4}{2}=6$ distinct pairs of simultaneously spiking neurons, and the
total number of distinct pairs is $\tbinom{5}{2}=10$. The normalized number
of spiking pairs is therefore $6/10$. Note that the first $m$ factorial
moments linearly determine the first $m$ power moments and vice versa,
because $\tbinom{n\ya}{m}$ is a polynomial in $\ya$ of degree $m$; so
fixing the ones is equivalent to fixing the others. Now we have this
extremely convenient property: \emph{the first $n$ normalized factorial
  moments for a sample and for the full network are numerically identical}:
\begin{equation}
  \label{eq:equal_m_factorialmoments}
  \E\condt[\big]{\tbinom{n\ya}{m}\;\tbinom{n}{m}^{-1}}{\yH} =
  \E\condt[\big]{\tbinom{N\yA}{m}\;\tbinom{N}{m}^{-1}}{\yH},
  \qquad m \le n.
\end{equation}

We can therefore apply the maximum-entropy method to obtain a distribution
for the full network, by constraining its $m$th factorial moment to be
equal to the sample's recorded average of
$\tbinom{n\ya}{m}\;\tbinom{n}{m}^{-1}$, for as many $m$ as we please with
$1\le m \le n$. The constraints on the distribution $\pf(\yA \| \yH)$ for
the full network are
\begin{equation}
  \label{eq:final_constraint}
  \frac{1}{T}\sum_{t}\binom{n\ya_{t}}{m}\,\binom{n}{m}^{-1}
  =
   \sum_{\yA}
  \binom{N \yA}{m}\,  \binom{N}{m}^{-1}\, \pf(\yA \| \yH)
\end{equation}
for any $m$s we wish.

@@

The number of moments used with this method depends on the specific
questions and hypotheses that a researcher is exploring; for example,
hypotheses of sufficient statistics, such as the sufficiency of pairwise
correlations for inferential purposes. In the present work we only want to
present the general method without entering into biological questions of
this kind. The particular case in which $n$ moments are constrained
corresponds to fully specify the measured time-frequency distribution of
the sample activity. In this case, our inferences about the full-network
activity are using all available measured frequency data. Using the full
frequency distribution of the sample may be a bad idea, however, because
the maximum-entropy distribution becomes a bad approximation of some of the
four distributions described above. It is preferable to use a moderately
high number of moments smaller than $n$. We explain this point in \sect***.

@@

We apply the approach just described to two data sets publicly available in
the literature:

The first data set, from \textcite[rat 14147]{stensolaetal2012}, consists
of $n=65$ neurons (27 of which classified as grid cells) from rat Medial
Entorhinal Cortex, recorded for about 20 minutes. Their spikes are
binned into $T=417\,641$ bins of $3\;\textrm{ms}$ width.

The second data set, from
\textcite{rostamietal2016_r2017}, %\texttt{data\_RS.npy}
consists of $n=159$ neurons from macaque Motor Cortex, recorded for about
15 minutes. Their spikes are binned into $T=300\,394$ bins of
$3\;\textrm{ms}$ width.

>>>>

\pdfoutput=1
%% Author: PGL  Porta Mana
%% Created: 2015-05-01T20:53:34+0200
%% Last-Updated: 2019-06-18T15:05:04+0200
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newif\ifarxiv
\arxivfalse
\ifarxiv\pdfmapfile{+classico.map}\fi
\newif\ifafour
\afourfalse% true = A4, false = A5
\newif\iftypodisclaim % typographical disclaim on the side
\typodisclaimtrue
\newcommand*{\memfontfamily}{zplx}
\newcommand*{\memfontpack}{newpxtext}
\documentclass[\ifafour a4paper,12pt,\else a5paper,10pt,\fi%extrafontsizes,%
onecolumn,oneside,article,%french,italian,german,swedish,latin,
british%
]{memoir}
\newcommand*{\updated}{\today}
\newcommand*{\firstdraft}{4 November 2015}
\newcommand*{\firstpublished}{***}
\newcommand*{\propertitle}{Maximum-entropy distributions for a 
  neuronal network\\ from subnetwork data%\\{\large ***}%
}
\newcommand*{\pdftitle}{\propertitle}
\newcommand*{\headtitle}{Maximum-entropy from sample data}
\newcommand*{\pdfauthor}{P.G.L.  Porta Mana, Y. Roudi, V. Rostami, E. Torre}
\newcommand*{\headauthor}{Porta Mana \etal}
\newcommand*{\reporthead}{}% Report number

@@

\newcommand{\appropto}{\mathrel{\vcenter{
  \offinterlineskip\halign{\hfil$##$\cr
    \propto\cr\noalign{\kern2pt}\sim\cr\noalign{\kern-2pt}}}}}
% https://tex.stackexchange.com/a/33547/97039
%\newcommand*{\fm}{f}
\newcommand*{\px}{p}

@@

We apply our approach to two concrete data sets: the activity of 65 neurons
recorded from a rat's Medial Entorhinal Cortex \mynote{ref}, and the
activity of 159 neurons recorded from a macaque's Motor Cortex
\mynote{ref}\mynote{add recording length}. In the first data set, the most
interesting inference is that the most probable frequency distribution for
the total activity of the full network, if the latter is assumed to be
larger than 1\,000 neurons, has two very distinct modes, both at low
activities ($0.7\%$ and $2.6\%$ of simultaneously active neurons within
$3\;\mathrm{ms}$, see \fig***). The analogous distribution for second data
set doesn't have two modes but still presents one prominent shoulder in its
one low-activity mode. Note that these inferred features for the full
network aren't observed in the sample, nor are they inferred by the
application of maximum-entropy to the sample alone. \mynote{if we manage
  the Monte Carlo sampling or other approx: say something about the
  deviation from the most probable frequency distribution}

@@

Our questions concern general features of the total activity $\yA$ of the
full network during and after the recording. For example: what was its
frequency distribution in time during the recording? How much total
activity should we expect at a time bin during the recording? And at a new
time bin? We cannot answer these questions with certainty; we can only give
distributions of probability over their possible answers. Our approach
gives such probability distributions.

@@

The idea behind the approach is easily summarized:
\begin{enumerate}[label=(\alph*)]
  \item\label{item:maxent_step} we build a distribution for the total
  activity of the full network using the maximum-entropy method;
  \item\label{item:sample_step} the constrained averages used in the
  maximum-entropy method for the full network are, in turn, determined via
  sampling theory from the constrained averages for the sample.
\end{enumerate}
Let's discuss these points in detail.

Regarding~\ref{item:maxent_step}, we assume that you're familiar with the
maximum-entropy method. In short, it amounts to a pair of prescriptions:
choose the distribution having maximum Shannon entropy among those
satisfying specific convex constraints, such as fixed averages; and judge
these averages to be equal to measured averages. We add two remarks about
this method that are seldom made in the literature. First, the
\enquote{distribution} given by this method, which we denote by $\px(\yA)$,
is the zeroth-order approximation of four different distributions for the
full network:
\begin{itemize}
\item the most probable \emph{frequency} distribution for the total
  activity across the \emph{recorded} bins, %$\yFF(\yA)$,
\item the most probable \emph{frequency} distribution for the total
  activity in a very long run of \emph{new} time bins, $\ynuu(\yA)$,
\item the \emph{probability} distribution for the value of the total
  activity at any time bin among those \emph{recorded},
  %$\yp(\yA_{t} \| \yH)$ with $1\le t \le T$,
\item the \emph{probability} distribution for the value of the total
  activity at a \emph{new} time bin.
  % $\yp(\yA_{t} \| \yH)$ with $t<1$ or $t>T$
\end{itemize}
The maximum-entropy distribution is thus an approximate answer to four
completely different inferences. Note that although the four distributions
above numerically coincide in this approximation, they differ in
higher-order approximations. We discuss their differences in \sect***.
Second, the maximum-entropy method based on the Shannon entropy implicitly
makes specific assumptions about the probabilities for the long-run
frequency distributions
\citep{jaynes1986d_r1996,portamana2009,portamana2017}. We discuss these
assumptions in \sect***.

In our case, to apply the method for the total activity of the full network
we need to fix some of the latter's averages, for example its moments. But
we don't have any measured moments for the full network to equate the
distribution moments to. Here enters point~\ref{item:sample_step}: the
probability calculus gives an exact, linear relation between the first $m$
moments for the full network and the first $m$ for the sample
\citep[\eqns~(16)]{portamanaetal2015}; the ones determine the others and
vice versa at every time bin. This relation is a classical result of
sampling theory
\citep[\chaps~I--IV]{whitworth1867_r1965}[\chap~II]{feller1950_r1968}[\chap~3]{jaynes1994_r2003}[see
also][]{whitworth1897}.

Combining this result with the maximum-entropy prescription
\enquote{moments${}={}$measured moments} for the sample, we have that the
measured moments for the sample determine the moments for the full network:
\begin{equation*}
  \label{eq:chain_prescription_moments}
  \mathrlap{\overbracket{\phantom{\text{measured moments} \rightarrow
        \text{sample moments}}}^{\textit{maximum-entropy prescription}}}
  \text{measured moments} \rightarrow
        \underbracket{\text{sample moments}
  \rightarrow
    \text{full-network moments}}_{\textit{sampling theory}}
\end{equation*}

These two steps are more straightforward if instead of power moments we use
\emph{normalized factorial moments} \citep{potts1953}. The $m$th normalized
factorial moment of a distribution $\px(\ya)$ for the activity of the
sample neurons is defined as the average
\begin{equation}
  % \E\condt[\Bigg]{
  %   \binom{n \ya}{m}\,\binom{n}{m}^{-1}}{\yH}
  % \equiv
  \sum_{\ya} %m!\,
  \binom{n \ya}{m}\,  \binom{n}{m}^{-1}\, \px(\ya),
  \qquad 1\le m \le n
\end{equation}
% where $\px(\ya \| \yH)$ is the probability for activity $\ya$ given
% assumptions $\yH$; \enquote{$\ya$} may refer to any recorded bin or to a
% new bin.
This moment can be interpreted as the expectation of the number of distinct
$m$-tuples of simultaneously spiking neurons (within a bin's time width),
normalized by the number of distinct $m$-tuples. For example, with $m=2$,
if $na=4$ neurons spike in a network of $n=5$, we have $\tbinom{4}{2}=6$
distinct pairs of simultaneously spiking neurons, and the total number of
distinct pairs is $\tbinom{5}{2}=10$. The normalized number of spiking
pairs is therefore $6/10$. Note that the first $m$ factorial moments
linearly determine the first $m$ power moments and vice versa, because
$\tbinom{n\ya}{m}$ is a polynomial in $\ya$ of degree $m$; so fixing the
ones is equivalent to fixing the others. Now we have this extremely
convenient property: \emph{the first $n$ normalized factorial moments for a
  sample and for the full network are numerically identical}:
\begin{equation}
  \label{eq:equal_m_factorialmoments}
    \sum_{\ya} %m!\,
  \binom{n \ya}{m}\,  \binom{n}{m}^{-1}\, \px(\ya)
=    \sum_{\yA} %m!\,
  \binom{N \yA}{m}\,  \binom{N}{m}^{-1}\, \px(\yA),
  % \E\condt[\big]{\tbinom{n\ya}{m}\;\tbinom{n}{m}^{-1}}{\yH} =
  % \E\condt[\big]{\tbinom{N\yA}{m}\;\tbinom{N}{m}^{-1}}{\yH},
  \qquad 1\le m \le n,
\end{equation}
where $\px(\yA)$ is the distribution for the full-network activity.

We can therefore apply the maximum-entropy method to obtain a distribution
for the full network, by constraining its $m$th factorial moment to be
equal to the sample's recorded average of
$\tbinom{n\ya}{m}\;\tbinom{n}{m}^{-1}$, for as many $m$ as we please with
$1\le m \le n$. The constraints on the distribution $\px(\yA)$ for
the full network are
\begin{equation}
  \label{eq:final_constraint}
  \frac{1}{T}\sum_{t}\binom{n\ya_{t}}{m}\,\binom{n}{m}^{-1}
  =
   \sum_{\yA}
  \binom{N \yA}{m}\,  \binom{N}{m}^{-1}\, \pf(\yA)
\end{equation}
for any $m$s we wish.

@@

The number of moments used with this method depends on the questions and
hypotheses that a researcher is exploring; for example, hypotheses of
sufficient statistics, such as the sufficiency of pairwise correlations for
inferential purposes. In the present work we only want to present the
general method without entering into biological questions of this kind.

The particular case in which $n$ moments are constrained is especially
important: it corresponds to fully specify the measured frequency
distribution for the activity of the sample neurons. In this case, our
inferences about the full-network activity are using all available measured
frequency data. Note that application of the maximum-entropy method
\emph{at the sample level} is in this case trivial and meaningless -- it
just gives back the measured frequency distribution. But application of the
method \emph{at the level of the full network}, as done here, is not
trivial.

Using the full frequency distribution of the sample may be a bad idea,
however, because the maximum-entropy distribution becomes a bad
approximation of some of the four distributions described above. It is
preferable to use a moderately high number of moments smaller than $n$. We
explain this point in \sect***.

@@

We apply the approach just described to two data sets publicly available in
the literature:
\begin{itemize}[wide]
\item The first data set, from \textcite[rat 14147]{stensolaetal2012},
  consists of $n=65$ neurons (27 of which classified as grid cells) from
  rat Medial Entorhinal Cortex, recorded for about 20 minutes. Their spikes
  are binned into $T=417\,641$ bins of $3\;\textrm{ms}$ width.
\item The second data set, from
  \textcite{rostamietal2016_r2017}, %\texttt{data\_RS.npy}
  consists of $n=159$ neurons from macaque Motor Cortex, recorded for about
  15 minutes. Their spikes are binned into $T=300\,394$ bins of
  $3\;\textrm{ms}$ width.
\end{itemize}
